---
title: "Getting Started with pslongSim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pslongSim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
set.seed(42)
```

## Overview

**pslongSim** simulates longitudinal clinical trial data with
treatment-driven side effects, an AR(1) biomarker, time-varying
adherence, and one of three outcome types (continuous, binary, or
time-to-event).  It provides a complete pipeline from data generation
through propensity score estimation, IPTW weighting, and marginal
structural model fitting.

## Quick Start

```{r}
library(pslongSim)

# Generate 200 patients across 3 treatment arms
demo <- generate_patient_data_demographic(num_patients = 200, seed = 1)
trt_map <- suppressWarnings(define_treatment_map(levels(demo$Treatment)))

# Simulate continuous outcomes with 6 visits
dat <- generate_longitudinal_data(
  demo, num_visits = 6, trt_map = trt_map,
  adherence_type = "beta", outcome_type = "continuous"
)

cat("Rows:", nrow(dat), "\n")
cat("Columns:", paste(names(dat), collapse = ", "), "\n")
```

## Propensity Score Estimation

Fit a propensity score model using GLM, Random Forest (ranger), or
XGBoost:

```{r}
ps_glm <- fit_ps_tidymodels(dat, model = "glm")
cat("PS range:", range(ps_glm$ps_hat), "\n")
```

## IPTW and MSM

Compute stabilised inverse probability of treatment weights and fit a
marginal structural model:

```{r}
dat$ps_hat <- ps_glm$ps_hat
dat_w <- compute_stabilized_iptw(dat)
cat("Weight range:", range(dat_w$w), "\n")

msm <- fit_msm_gee_cont(dat_w)
summary(msm)
```

## Binary Outcome with GEE MSM

For binary outcomes, use `outcome_type = "binary"` and
`fit_msm_gee_bin()`:

```{r}
dat_bin <- generate_longitudinal_data(
  demo, num_visits = 6, trt_map = trt_map,
  adherence_type = "binary", outcome_type = "binary"
)
ps_bin <- fit_ps_tidymodels(dat_bin, model = "glm")
dat_bin$ps_hat <- ps_bin$ps_hat
dat_bin_w <- compute_stabilized_iptw(dat_bin)
msm_bin <- fit_msm_gee_bin(dat_bin_w)
summary(msm_bin)
```

## Time-to-Event with Cox MSM

For time-to-event outcomes, use `outcome_type = "tte"` and
`fit_msm_cox()`:

```{r}
tte <- generate_longitudinal_data(
  demo, num_visits = 8, trt_map = trt_map,
  adherence_type = "binary", outcome_type = "tte"
)
ivl <- filter_at_risk(tte$tte_intervals)
ps_tte <- fit_ps_tidymodels(ivl, model = "glm")
ivl$ps_hat <- ps_tte$ps_hat
ivl_w <- compute_stabilized_iptw(ivl)
cox_m <- fit_msm_cox(ivl_w)
summary(cox_m)
```

## Random Forest Propensity Scores

Use `model = "ranger"` for random forest PS estimation (requires the
`ranger` package):

```{r eval=requireNamespace("ranger", quietly=TRUE)}
ps_rf <- fit_ps_tidymodels(dat, model = "ranger")
cat("RF PS range:", range(ps_rf$ps_hat), "\n")
```

## Non-Linear Feature Detection

One of pslongSim's key features is the ability to generate data with
non-linear interactions that are **undetectable by linear regression**
but **readily detectable by tree-based models**.

When `non_linear_feature = TRUE`, interaction terms (e.g.,
`Age * SideEffect`) are added to the outcome with no additive main
effects.  Linear models can only capture additive effects, so they
miss this signal entirely.  Random forests and XGBoost, which
naturally partition on feature interactions, detect it easily.

```{r}
# Generate data WITH the non-linear interaction
set.seed(42)
demo <- generate_patient_data_demographic(num_patients = 1000, seed = 1)
trt_map <- suppressWarnings(define_treatment_map(levels(demo$Treatment)))

dat_nl <- generate_longitudinal_data(
  demo, num_visits = 6, trt_map = trt_map,
  adherence_type = "beta", outcome_type = "continuous",
  non_linear_feature = TRUE,
  nonlin_coefs = c("Age:SideEffect" = 0.20)
)

cat("NL columns:", paste(grep("^NL_", names(dat_nl), value = TRUE), collapse = ", "), "\n")
```

### Linear Regression vs Random Forest

```{r rf-comparison, eval=requireNamespace("randomForest", quietly=TRUE)}
# Fit a linear model (additive only — cannot capture the interaction)
lm_additive <- lm(
  Outcome ~ Treatment + VisitNumber + Age + GenderMale +
    SideEffect + Biomarker + Compliant,
  data = dat_nl
)

# Fit a random forest
rf_fit <- randomForest::randomForest(
  Outcome ~ Treatment + VisitNumber + Age + GenderMale +
    SideEffect + Biomarker + Compliant,
  data = dat_nl, ntree = 200
)

cat("Linear model R-squared:", round(summary(lm_additive)$r.squared, 4), "\n")
cat("Random Forest R-squared:", round(1 - rf_fit$mse[200] / var(dat_nl$Outcome), 4), "\n")
```

The random forest R-squared substantially exceeds the linear model
R-squared because the RF captures the `Age * SideEffect` interaction
that the linear model misses entirely.

### Custom Interactions

You can specify multiple interactions via `nonlin_coefs`:

```{r}
dat_multi <- generate_longitudinal_data(
  demo, num_visits = 6, trt_map = trt_map,
  adherence_type = "beta", outcome_type = "continuous",
  non_linear_feature = TRUE,
  nonlin_coefs = c(
    "Age:SideEffect"   = 0.15,
    "Biomarker:Age"    = 0.10
  )
)

cat("NL columns:", paste(grep("^NL_", names(dat_multi), value = TRUE), collapse = ", "), "\n")
```

## Using the Pipeline

The full simulation workflow is packaged as a `targets` pipeline:

```{r eval=FALSE}
library(pslongSim)
# Requires the 'targets' package
targets::tar_make()

# Read results
targets::tar_read(summary_cont)
targets::tar_read(msm_bin)
targets::tar_read(summary_cox)
```

## Session Info

```{r}
sessionInfo()
```
