---
title: "Longitudinal Propensity Score Analysis: Method Comparison"
geometry: margin=1in
author: 
- name: J Kyle Armstrong, PhD
date: today
format:
  pdf:
    toc: true
    number-sections: true
    pdf-engine: pdflatex
---

```{r}
#| label: setup
#| echo: false
#| message: false

library(targets)
library(tidyverse)
library(knitr)
library(MLPScore)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

# Load targets
# Load targets dynamically depending on working directory
if (dir.exists("../../_targets")) {
    tar_config_set(store = "../../_targets")
} else {
    tar_config_set(store = "../_targets")
}
tar_load(base_params)
tar_load(all_results)
tar_load(fig1)
tar_load(fig2)
tar_load(fig3)
tar_load(fig4)
tar_load(fig5)
tar_load(fig6)
tar_load(fig7)

# Prepare true effects for plotting
# Derive Term names dynamically (e.g., TreatmentTreatment_2) to match model output
rterm_levels <- unique(all_results$Term)
true_effects_df <- data.frame(
    Term = rterm_levels,
    True_Effect = sapply(rterm_levels, function(x) {
        if (grepl("Treatment_2", x)) {
            return(as.numeric(base_params$trt2_outcome_effect))
        }
        if (grepl("Treatment_3", x)) {
            return(as.numeric(base_params$trt3_outcome_effect))
        }
        return(0)
    })
)
```

## Objective

This report compares treatment effect estimates across various analytic strategies and scenarios of confounding ($\\rho_{uv}$). We explore variations of the populations (Intention-to-Treat (ITT), Per-Protocol (PP), and Propensity Score Matching (PSM)) across both visit-level and patient-level adherence.

### Outcome Type Interpretations

The simulation framework dynamically adapts to the selected outcome type (`r if (!is.null(base_params$outcome_type)) base_params$outcome_type else "continuous"`). Depending on the configuration, the results in this report should be interpreted as follows:

*   **Continuous Outcomes:** Treatment effects are modeled using standard linear/mixed-effects models. The `Estimate` (and Bias) represents the absolute difference in expected means between the treatments.
*   **Binary Outcomes:** Treatment effects are modeled using logistic regression with binomial link functions. The `Estimate` (and Bias) is interpreted on the log-odds scale, and the estimates represent log-odds ratios.

\clearpage
## Representative Scenario (Rho=0.5, Adherence ~75%)

### Patient Level Comparisons
```{r fig-1-2}
fig1
fig2
```

### Observation Level Comparisons
```{r fig-3}
fig3
```

\clearpage
## Method Comparison (Pooled)

Below we compare the estimated treatment effects across all scenarios.

```{r analysis}
all_results %>%
    group_by(Level, Method, Term) %>%
    summarize(
        Mean_Estimate = mean(Estimate),
        SD_Estimate = sd(Estimate),
        .groups = "drop"
    ) %>%
    arrange(Term, Level, Method) %>%
    kable(caption = "Averaged Treatment Effects Across Scenarios (including Twang)", digits = 3)
```

\clearpage
## Visualizing Bias and Coverage

### Bias vs Confounding Strength (Rho Sweep)
```{r fig-4-5}
fig4
fig5
```

### Bias vs Observed Adherence Rate (Adherence Sweep)
```{r fig-6-7}
fig6
fig7
```

\clearpage
## Longitudinal Estimates Distribution

Grouping by $\rho_{uv}$ to see how estimates distribute relative to truth.

```{r plot-rho}
all_results %>%
    left_join(true_effects_df, by = "Term") %>%
    ggplot(aes(x = as.factor(rho_uv), y = Estimate, color = Method)) +
    geom_boxplot() +
    facet_wrap(~ Level + Term, scales = "free_y") +
    geom_hline(aes(yintercept = True_Effect), linetype = "dashed", alpha = 0.5) +
    labs(
        title = "Treatment Effect Estimates by Confounding Strength",
        x = "Rho_uv (Confounding Strength)",
        y = "Estimate"
    ) +
    theme_minimal()
```

\clearpage
## Discussion

Machine learning PS methods (GBM, XGBoost, RF) along with `twang` are evaluated against parametric GLM and naive ITT/PP approaches. As $\\rho_{uv}$ increases, the selection bias in PP approaches typically grows, while PS methods attempt to recover the true effect. The analysis confirms that understanding the adherence-outcome correlation is critical in deciding the appropriate analytic approach, especially in the context of the specific outcome type evaluated.
